Validation of marker gene expression

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
new template for validation reports

---
title: "Validation of marker gene expression"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000001"
---

@@

# results directory with cell-type-consensus 
results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)

# data directory where project metadata files live
data_dir <- file.path(repository_base, "data", "current")
```

@@

```{r}
# get consensus results 
consensus_results_files <- list.files(results_dir, pattern = "_consensus-cell-type-assignments\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)

# get gene expression files 
gene_exp_files <- list.files(results_dir, pattern = "_marker-gene-expression\\.tsv.\\gz$", recursive = TRUE, full.names = TRUE)

# list of all validation markers
validation_markers_file <- file.path(module_base, "references", "validation-markers.tsv")

@@

```{r}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
combined_df <- consensus_df |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    detected = logcounts > 0 # column indicating if gene is present or not
  ) 
```

@@

@@

Note that I did try and look at all marker genes, without specifying that they must be unique, and the plots were overwhelming with lots of genes. 
Instead I included the a separate section looking at the mean expression across all markers for a given group.

### Dot plot

@@

### Heatmap

The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
In contrast to the dot plot, we will show all marker genes for a given cell type group, not just the uniquely expressed genes.

```{r}
# add column to uniquely label gene symbols since some genes are present across multiple cell types 
# add this here since we will use the order in this data frame to set the order of the heatmap columns
heatmap_markers_df <- filtered_markers_df |> 
  dplyr::mutate(
    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
  ) 
```

```{r}
# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- group_stats_df |> 
  dplyr::ungroup() |> 
  # add unique label for gene symbol 
  dplyr::mutate(
    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
  ) |> 
  dplyr::select(unique_symbol, broad_celltype_group, mean_exp) |> 
  dplyr::filter(unique_symbol %in% heatmap_markers_df$unique_symbol) |> 
  tidyr::pivot_wider(
    names_from = unique_symbol,
    values_from = mean_exp
  ) |> 
  tibble::column_to_rownames("broad_celltype_group") |>
  as.matrix()
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[c(celltype_order, "unknown"), heatmap_markers_df$unique_symbol]

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  group = heatmap_markers_df$validation_group_annotation,
  col = list(
    group = celltype_colors
  ),
  annotation_legend_param = list(
      title = "Validation group"
    )
)
```

```{r, fig.height=7, fig.width=10}
marker_gene_heatmap(heatmap_mtx, annotation)
```

@@

## Conclusions

- Regardless of looking at individual genes or the mean expression of all genes, we see similar patterns where generally the cells assigned to a given cell type show expression of the appropriate marker genes when compared to other cells. 
- There seems to be pretty clear distinction of T cells, B cells, and NK cells. 
- Monocytes, macrophages, and myeloid cells are generally hard to distinguish from each other and all share similar marker genes. 
Perhaps we could collapse these into myeloid, but I do see expression of monocyte specific genes in monocytes only and maybe some macrophage specific genes in macrophages only, so it's probably okay to keep separate. 
I think that is more obvious in the heatmaps than in the dot plots. 
- Neurons and astrocytes both show expression of both neuron and astrocyte marker genes. 
It does look like there are a set of genes that are more highly expressed in the astrocyte population though, so I think we can keep those gene sets separate. 
- Here the unknown cells don't show strong expression of any specific set of genes, but it does look like they show more expression of neuron/astrocyte markers than any of the other marker genes in the plot. 
This makes sense since we are looking at brain tumors and I expect similarity between tumor cells and neural cell types.

>>>>

---
title: "Validation of marker gene expression"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000002"
---

@@

# path to results directory
results_dir <- file.path(repository_base, "data", "current", "results", "cell-type-consensus", params$project_id)
```

@@

```{r}
# get consensus results 
consensus_results_files <- list.files(results_dir, pattern = "_processed_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

# get gene expression files 
gene_exp_files <- list.files(results_dir, pattern = "_processed_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

# check that consensus results exist
stopifnot(
  "No cell type results found in data/current/results/cell-type-consensus" = length(consensus_results_files) > 0,
  "No gene expression files found in data/current/results/cell-type-consensus" = length(gene_exp_files) > 0
)
```

```{r}
# list of all validation markers
validation_markers_file <- file.path(module_base, "references", "validation-markers.tsv")

@@

```{r}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
combined_df <- consensus_df |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    detected = logcounts > 0 # column indicating if gene is present or not
  )

# remove the large data frames 
rm(consensus_df)
rm(gene_exp_df)
```

@@

# no longer need this and it takes up space 
rm(combined_df)

@@

@@

@@
